\chapter{Extending A Genetic Algorithm Model To The Diploid Case} \label{ch:GA model Diploid}

This chapter describes a simple Markov model for evolution under the
influence of crossing over and mutation; it is a non-overlapping,
generational, infinite population model under the assumption of {\em complete panmixia} (random mating) and no
selective pressure. This chapter shows how diploid evolution equations can be represented by haploid equations 
and can be specialized to Vose's infinite population model, which is a haploid model.

A basic syntactic model for haploid and diploid genomes is first considered. 
Then the mechanics of how next generation is obtained from current generation are
defined abstractly in procedural terms, which serves to motivate the equations governing evolution. 
Next evolution equations are developed corresponding to the
procedural description defining evolution for a population of
diploid genomes. Observations concerning the form and symmetry of
those equations directly lead to decoupling from the diploid case a
haploid model sufficient to determine evolutionary trajectories for
the diploid case. Mask based mutation and crossover operators are used to specialize 
haploid equations to Vose's infinite haploid population model. 
Analytical and computational simplification resulting from specialization to 
Vose's infinite population model are explained and used 
in experimental simulation and study of convergence of finite population 
short-term behavior to behavior predicted by infinite population model.    

\section{Model} \label{Model}
A haploid genome $g$ is defined syntactically as a length $\ell$
binary string.  A collection of $h$ chromosomes may be modeled by
partitioning $g$ into $h$ segments (of arbitrary lengths $\ell_1,
\ldots , \ell_h$; thus $\ell = \ell_1 + \cdots + \ell_h$).

A diploid genome $\alpha = \langle \alpha_0, \alpha_1 \rangle$ is
likewise defined syntactically as a pair of length $\ell$ binary
strings.  Although simple, that syntax is flexible and possesses
significant modeling power by means of tailoring partitioning to
application.  We concentrate on the abstract level, considering the
evolution of a non-overlapping, generational, infinite population
model assuming panmixia and no selective pressure. We are not concerned with 
whether and how partitioning is defined as it is irrelevant 
to the development.

Following Hardy (see \cite{Hardy1908}), the model $q^{n}$ at generation $n$
is a vector having for component $q_\alpha^n$ the prevalence of
diploid $\alpha\,$ (the probability of selecting $\alpha$ \nudge at
generation $n$, assuming unbiased selection).\footnote{The
  representation here is the conceptual equivalent of Hardy's model.}
Ordered diploid $\gamma = \langle \gamma_0, \gamma_1 \rangle$ is
produced for generation $n+1$ according to following procedural
description.

  Assuming independent selection events:
\begin{itemize}
\item From parent $\alpha$ --- selected with probability
  $q_\alpha^n$ --- obtain gamete $\gamma_0$
\item From parent $\beta$ --- selected with probability $q_\beta^n$
  --- obtain gamete $\gamma_1$
\end{itemize}
Following Gieringer (see \cite{Geiringer1944}), let the transmission
function $t_\alpha(g)$ be the probability that gamete $g$ is produced
from parental genome $\alpha$.  It follows from the above that the
equation determining the next generation $q^{n+1}$ is
\begin{equation}
\label{model0}
q_\gamma^{n+1} \; = \;
\sum_{\alpha} \, q_\alpha^n \, t_\alpha(\gamma_0) 
\sum_{\beta} \,q_\beta^n \, t_\beta(\gamma_1)\\[-.05in]
\end{equation}

It should be appreciated that the Mendelian (see \cite{Mendel1866}) laws of
segregation\footnote{Alleles of a given locus segregate into separate
  gametes.} and independent assortment\footnote{Alleles of one gene
  sort into gametes independently of the alleles of another gene.}
need not be respected by the transmission function.


The right hand side of (\ref{model0}) is invariant under interchange
of the summation variables $\alpha$ and $\beta$, which is equivalent
to interchanging $\gamma_0$ and $\gamma_1$.  This symmetry reflects
the fact that which haploid of $\gamma$ is designated as $\gamma_0$ is
arbitrary,\\[-.05in]
\[
q_{\langle \gamma_0, \gamma_1 \rangle}^{n+1} \; = \;
q_{\langle \gamma_1, \gamma_0 \rangle}^{n+1}\\[.1in]
\]
The model corresponding to (\ref{model0}) is low-level in the sense
that it regards $\langle \gamma_0, \gamma_1 \rangle$ and $\langle
\gamma_1, \gamma_0 \rangle$ as distinct when $\gamma_1 \neq \gamma_0$.
A higher-level model based on sets is easily obtained,
\[
q_{\{\gamma_0, \nudge \gamma_1\}} \; = \; \left\{
\begin{array}{ll}
2 \nudge q_{\langle \gamma_0, \gamma_1 \rangle} & \mbox{ if $\gamma_0 \neq \gamma_1$}\\
\phantom{2 \nudge }q_{\langle \gamma_0, \gamma_1 \rangle} & \mbox{ otherwise}
\end{array}
\right.\\[.05in]
\]
which is in agreement with Hardy (see \cite{Hardy1908}).

\section{Reduction}

Evolution equation (\ref{model0}) may be reduced to the haploid case.
Its right hand side is the product of two summations; denote the first
by $p_{\gamma_0}^{n+1}$ and the second by $p_{\gamma_1}^{n+1}$ so that
\begin{equation}
\label{model1}
q_{\langle \gamma_0, \gamma_1 \rangle}^{n+1} \; = \;
p_{\gamma_0}^{n+1} \, p_{\gamma_1}^{n+1}\\[.05in]
\end{equation}
where for any haploid $\gamma_0$,
\begin{equation}
\label{model00}
p_{\gamma_0}^{n+1} \; = \;
\sum_{\alpha} \,q_\alpha^n \, t_\alpha(\gamma_0)
\end{equation}
It suffices to determine the evolution of the distributions $p^{n}$.
Uncoupling \nudge $p$ \nudge from \nudge $q$ \nudge using
(\ref{model00}), and equation (\ref{model1}) with superscript $n$ ---
instantiate the $n$ in (\ref{model1}) with $n-1$ --- yields the
evolution equation
\begin{eqnarray}
\label{model2}
p_{\gamma_0}^{n+1} & = &
\sum_{\alpha_0, \, \alpha_1} \, q_{\langle \alpha_0, \,\alpha_1 \rangle}^n \,
t_{\langle \alpha_0, \,\alpha_1 \rangle}(\gamma_0) \nonumber \\
& = &
\sum_{\alpha_0, \, \alpha_1} \, p_{\alpha_0}^n \, p_{\alpha_1}^n \,
t_{\langle \alpha_0, \,\alpha_1 \rangle}(\gamma_0) 
\end{eqnarray}
The $p^n$ are in fact distributions; summing equation
(\ref{model1}) with superscript $n$ yields
\[
1 \; = \; \sum_\alpha \, q_\alpha^n \; = \;
\sum_{\alpha_0, \, \alpha_1} \, p_{\alpha_0}^n \, p_{\alpha_1}^n \; = \;
\Big( \sum_{\alpha_0} \, p_{\alpha_0}^n \Big)^2
\]
Let $[\mbox{\em expression\/} ]$ denote $1$ if {\em expression\/} is
true, and $0$ otherwise.\footnote{$[ \cdots ]$ is sometimes referred to
  as an {\em Iverson bracket}.}  The weighted count of haploid
$g$ in generation $n$ is
\begin{eqnarray}
\label{project}
  & &
  \sum_{\alpha_0, \, \alpha_1} \, q_{\langle \alpha_0, \alpha_1 \rangle}^n
([g = \alpha_0] + [g = \alpha_1]) \\ & = &
\sum_{\alpha_0, \, \alpha_1} \, p_{\alpha_0}^n \, p_{\alpha_1}^n [g = \alpha_0] + 
\sum_{\alpha_0, \, \alpha_1} \, p_{\alpha_0}^n \, p_{\alpha_1}^n [g = \alpha_1] \\[0.05in]
& = & 2 \nudge p_g^n
\end{eqnarray}

Hence the (normalized) prevalence of haploid $g$ in generation $n$ is
the $g\,$th component of the distribution $p^n$. \linebreak
Moreover, (\ref{project}) and (\ref{model1}) show (for $n >
0$) invertibility of the map
\[
  \pi \nudge : \nudge {\bm q}^{n} \; \longmapsto \; {\bm p}^{n}
\]

Evolution equation (\ref{model2}) in matrix form is
\begin{equation}
\label{model3}
p_g^\prime \; = \; p^T M_g \,\nudge p
\end{equation}
where current state $p$ (generation $n$) and next state $p^\prime$
(generation $n+1$) are column vectors, and the $g\,$th transmission
matrix is
\begin{equation} \label{Mg}
\Big(M_g \Big)_{u,v} \; = \; t_{\langle u, v \rangle}(g)
\end{equation}
(vectors and matrices are indexed by haploids --- length $\ell$ binary
strings).

\input{chapters/specialization.tex}

\input{chapters/convergence.tex}

\section{Summary}
In this chapter, we began with a description of simple diploid Markov model under mutation 
and crossover with no selective pressure. With reduction to haploid and specialization 
using masked base recombination operators, we showed Vose's infinite population model, 
can be extended to diploid case. Using computational benefits 
of this reduction (to haploid model and Walsh transform), we showed via experiment and 
regression of resulting data that distance between finite diploid population and infinite 
diploid population decreases like $1/\sqrt{N}_{th}$ which is consistent with 
the single step-step haploid case convergence behavior predicted by 
Vose's infinite population model.






