\chapter*{Vita} \label{ch:vita}
Mahendra Duwal Shrestha was born in Kathmandu, Nepal, 
to the parents Krishna Prasad and Laxmi Duwal Shrestha. 
He is the first of three children: Ambika and Anuka. 
He attended Panga Secondary School up to tenth grade. 
Graduating tenth grade in distinction division, he attended 
NIC Higher Secondary School, where he was also awarded scholarship, taking Physics and Mathematics major. 
After graduation, he attended Pulchowk Engineering Campus, Institute of Engineering, Tribhuvan University 
where he was merit listed student for four years. He was introduced to applied mathematics, programming in 
C and C++, and electronic circuits and communication. 
He got bachelor's degree of Engineering in Electronics and Communication 
from Pulchowk Engineering Campus, IOE, Tribhuvan University in March 2011. 
He worked for two and half years as software engineer at Deerwalk Services Pvt. Ltd, 
developing health care business applications for clients in US. 
During the short stint at Deerwalk Services, he learnt and applied knowledge in programming and scripting languages as C#, 
Java, Groovy and javascript, web framworks as ASP.NET, ASP.NET MVC, Grails, Ajax and jquery, and databases as MYSQL and MSSQL. 
He accepted teaching assistantship at The University of Tennessee, Knoxville in Electrical Engineering and Computer Science 
program. He is continuing his education in Masters of Science in Computer Science working as research assistant at Sergey's lab 
in Ecology and Evolutionary Biology.
